A Mechanised Library for Generic (Co)Programming

### 2024-02-23 paper/paper.tex
Up to recursive hylos

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Definition rana_f__ `{equiv A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{minted}
Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
the Coq recursive definition. We then couple coalgebras with a proof that they
are recursive, and add a coercion from recursive coalgebras to coalgebras.
\begin{minted}{Coq}
Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
\end{minted}
Similarly to the rest of the definitions, we prove that recursive anamorphisms
are respectful, and use this fact to build them as higher order proper
morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
the unique solution to the equation 
\mintinline{Coq}{f =e l_in \o fmap f \o h},
where \mintinline{Coq}{h} is a recursive coalgebra.
\begin{minted}{Coq}
Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

@@

\subsection{Recursive Hylos}

Recursive hylomorphisms are built from an algebra and a recursive coalgebra:
\begin{minted}{Coq}
Definition hylo_def (a : Alg F B) (c : Coalg F A)
: forall (x : A), RecF c x -> B
:= fix f x H :=
  match c x as h0
        return
        (forall e : Pos (shape h0), RecF c (cont h0 e)) ->
        B
  with
  | MkCont s_x c_x => fun H => a (MkCont s_x (fun e => f (c_x e) (H e)))
  end (RecF_inv H).
\end{minted}

\begin{minted}{Coq}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{minted}

Recursive hylo universal property allows us to prove common optimisations:

\begin{minted}{Coq}
    Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) (g2 : Alg F C)
    (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
    : f2 \o hylo g1 h1 =e hylo g2 h1.
  Proof.
    rewrite hylo_univ.
    rewrite fmap_comp.
    rewrite compA.
    rewrite <- E2.
    rewrite <- compA.
    rewrite <- compA.
    rewrite (compA g1).
    rewrite <- hylo_unr.
    reflexivity.
  Qed.

  Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
    (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
    : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

  Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
    (g : Alg F B) (h : RCoalg F A)
  : cata g \o rana h =e hylo g h.
\end{minted}

>>>>

However, there still remains the issue of reasoning about the equality of
container extensions. The contents of the container extensions need to be
accessed via a function \mintinline{Coq}{contents}. The domain of this function
depends on the shape of the container. This means that, for a given type of
shapes and positions, the equality of two objects of a container extension will
need to deal with heterogeneous equality. Since we want to remain axiom-free,
one way to deal with this is to rely on a \emph{decidable} validity predicate.
This way, we reach our final definition for container extensions. First,
containers are defined as pairs of shape and position types that have a
\emph{decidable} validity predicate. We represent decidable predicates as
functions that return a boolean, and lift booleans to propositions similarly to
SSReflect. For simplicity, we define the dependent pair of valid positions as a
record, which will be extracted to a singleton type with the validity proof
erased by Coq's code extraction mechanism.
\begin{minted}{Coq}
Class Cont (Shape : Set) (Pos : Set) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) := 
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{minted}
Then, we define \emph{container extensions} as
\begin{minted}{Coq}
Record App `(C : Cont Shape Pos) (X : Set) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{minted}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{minted}{Coq}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{minted}

@@

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkApp sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

Definition rana_f__ `{equiv A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkApp (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{minted}
Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
the Coq recursive definition. We then couple coalgebras with a proof that they
are recursive, and add a coercion from recursive coalgebras to coalgebras.
\begin{minted}{Coq}
Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
\end{minted}
Similarly to the rest of the definitions, we prove that recursive anamorphisms
are respectful, and use this fact to build them as higher order proper
morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
the unique solution to the equation 
\mintinline{Coq}{f =e l_in \o fmap f \o h},
where \mintinline{Coq}{h} is a recursive coalgebra.
\begin{minted}{Coq}
Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

@@

\subsection{Recursive Hylomorphisms}

Given an $F$-algebra and $F$-coalgebra, a solution to the equation $f = a \circ
F\; f \circ c$ is known as a hylomorphism. Unfortunately, this definition
cannot be accepted by Coq, since there is no structurally smaller argument to
do recursion, and there is no corecursive definition that is guaranteed to be
guarded, due to the use of an arbitrary algebra $a$. Furthermore, solutions to
the hylomorphism equation are \emph{not} guaranteed to be unique, or to even
exist. To address these issues, we mechanise \textbf{recursive hylomorphisms}
which, as we will see, are guaranteed to have a unique solution.

Hylomorphisms \emph{are} the unique solution to $f = a \circ F\; f \circ c$, if
$c$ is a recursive coalgebra. We show this by first providing the definition
for hylomorphisms in Coq, and then proving that any function satisfying this
equation is a hylomorphism. We call these hylomorphisms with a recursive
coalgebra, \textbf{recursive hylomorphisms}.
\begin{minted}{Coq}
Definition hylo_def (a : Alg F B) (c : Coalg F A) 
  : forall (x : A), RecF c x -> B := fix f x H 
  := match c x as h0 return
       (forall e : Pos (shape h0), RecF c (cont h0 e)) -> B
     with
     | MkApp s_x c_x => fun H => a (MkApp s_x (fun e => f (c_x e) (H e)))
     end (RecF_inv H).
\end{minted}
The definition of recursive hylomorphisms is recursive on the proof that the
anamorphism can only be applied finitely many times on any input. We again use
\mintinline{Coq}{RecF_inv} to obtain the structurally smaller proof to use in
the recursive calls.  As we did with catamorphisms and recursive anamorphisms,
we prove that \mintinline{Coq}{hylo_def} is respectful, and use this proof to
built the corresponding higher-order proper morphism:
\begin{minted}{Coq}
Definition hylo : Alg F B ~> RCoalg F A ~> A ~> B := ...
\end{minted}
Finally, we show that recursive hylomorphisms are the unique solution to
the hylomorphism equation.
\begin{minted}{Coq}
Lemma hylo_univ (g : Alg F B) (h : RCoalg F A) (f : A ~> B)
    : f =e hylo g h <-> f =e g \o fmap f \o h.
\end{minted}
This result can be used to prove hylomorphism laws in a way that is analogous
to pen-and-paper program calculation techniques. We show the rule for 
hylomorphism fusion as an example:
\begin{minted}{Coq}
Lemma hylo_fusion_l (h1 : RCoalg F A) (g1 : Alg F B) 
  (g2 : Alg F C) (f2 : B ~> C) (E2 : f2 \o g1 =e g2 \o fmap f2)
  : f2 \o hylo g1 h1 =e hylo g2 h1.
Proof.
  (* f2 \o hylo g1 h1 =e hylo g2 h1 *)
  rewrite hylo_univ.
  (* f2 \o hylo g1 h1 =e (g2 \o fmap (f2 \o hylo g1 h1)) \o h1 *)
  rewrite fmap_comp.
  (* f2 \o hylo g1 h1 =e (g2 \o (fmap f2 \o fmap (hylo g1 h1))) \o h1 *)
  rewrite compA.
  (* f2 \o hylo g1 h1 =e ((g2 \o fmap f2) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- E2.
  (* f2 \o hylo g1 h1 =e ((f2 \o g1) \o fmap (hylo g1 h1)) \o h1 *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e (f2 \o g1) \o (fmap (hylo g1 h1) \o h1) *)
  rewrite <- compA.
  (* f2 \o hylo g1 h1 =e f2 \o (g1 \o (fmap (hylo g1 h1) \o h1)) *)
  rewrite (compA g1).
  (* f2 \o hylo g1 h1 =e f2 \o ((g1 \o (fmap (hylo g1 h1)) \o h1) *)
  rewrite <- hylo_unr.
  (* f2 \o hylo g1 h1 =e f2 \o hylo g1 h1 *)
  reflexivity.
Qed.
\end{minted}
Using the hylo fusion law, we can prove the well-known \emph{deforestation
optimisation}, where two consecutive recursive computations, one that builds a
data structure, and another one that consumes it, can be fused together into a
single recursive definition. This, in turn, allows us to prove that a recursive
hylomorphism is the composition of a catamorphism and a recursive anamorphism.
\begin{minted}{Coq}
Lemma deforest (h1 : RCoalg F A) (g2 : Alg F C)
  (g1 : Alg F B) (h2 : RCoalg F B) (INV: h2 \o g1 =e id)
  : hylo g2 h2 \o hylo g1 h1 =e hylo g2 h1.

Corollary cata_ana_hylo `(F : Cont Sh P) `{equiv A} `{equiv B}
  (g : Alg F B) (h : RCoalg F A)
: cata g \o rana h =e hylo g h.
\end{minted}
This result will allow us to \emph{calculate} programs by providing small, easy
to verify implementations, and combining them into efficient recursive
functions by \emph{using Coq's proofs of hylomorphism laws as optimisation
passes}.

\section{Extraction}

\dcas{Talk About this}
\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}
